---
title: "Preliminary results"
subtitle: Anova, Correlation, Pricincipal components, Least Discriminant analysis
author: 
  - name: Adamson Ahmed
    affiliation: Universitat Politecnica de Valencia
  - name: Gresia Eldin
    affiliation: Universitat Politecnica de Valencia
  - name: Ngwankfu Kelly 
    affiliation: Universitat Politecnica de Valencia
  - name: Miguel Hejase
    affiliation: Universitat Politecnica de Valencia
title-block-banner: true
format: 
  html:
    title-block-banner: true
---

## Title : Plant morphological traits and ecological stoichiometry in disturbed *vs*. conserved forests

**Data description** : Table S4 shows the data for various functional traits including morphological and nutrient content from leaves collected from disturbed and conserved forests.

**Link supplementary data** : <https://peerj.com/articles/20301/#supp-1>

**Variables** :

1\. Forest type (disturbed vs conserve)

2\. Plant species : 62 species in disturbed forests and 43 species in conserved forests

3\. Traits : 12 morphological traits and 24 ecological stoichiometry traits

4\. Left nutrients concentration and Ratio describing plant nutrient balance

## Introduction

Forests are globally significant ecosystems, providing critical ecosystem services such as carbon sequestration, climate regulation, soil protection, and the maintenance of essential biodiversity. The functional integrity of these ecosystems is intrinsically linked to the traits of the dominant plant communities that inhabit them. Understanding how plant traits vary in response to environmental filtering is paramount for effective conservation and management planning. Plant functional traits—specifically morphological traits (like leaf area or tissue density) and ecological stoichiometric traits (like carbon, nitrogen, and phosphorus content and ratios)—serve as measurable proxies for an organism's life history strategy and its resource economy (Source 12, 15). By comparing these traits between conserved and disturbed forest areas, researchers can identify shifts in community-wide resource allocation strategies. Trait variation often reflects a fundamental biological trade-off: a conservative strategy prioritizing longevity, structure, and low resource turnover (typically found in stable environments) versus an acquisitive strategy prioritizing rapid growth, high metabolic rates, and quick resource acquisition (often found in resource-rich or disturbed environments).

**Objectives** :

1.  To analyze and compare plant functional traits (morphological and stoichiometric traits) in disturbed and conserved forests.
2.  To examine relationships among various functional traits to investigate disparities in life history strategies between the two forest types.

**Research questions** :

1.  Is there any significant comparison in plant functional traits (morphological and stoichiometric traits) between disturbed and conserved forests.
2.  Is there significant correlation among various functional traits and the differences in life history strategies between the two forest types.

**Main hypotheses** :

1.  Null Hypothesis (HO): There is no significant difference in plant morphological traits and ecological stoichiometry between disturbed and conserved forests.

    Alternate Hypothesis (H1): Plant morphological traits and ecological stoichiometry differ significantly between disturbed and conserved forests.

2.  Null Hypothesis (HO): There are no correlation among various functional traits and no differences in life history strategies between the two forest types.

    Alternate Hypothesis (H1): There are significant correlation among various functional traits and significant differences in life history strategies between the two forest types.

**Main analyses :**

1.ANOVA 2.Correlation 3.Principal Components Analysis 4.Discriminant Analysis

## Methods

The methods employed in this study encompassed data acquisition, preparation, and a series of multivariate statistical analyses to compare plant functional trait syndromes between conserved and disturbed forest sites. All data preparation and statistical analyses were performed using the R statistical environment (Version 4.3.0, R Core Team). The dataset used for this project was obtained from supplementary material (specifically Table S4) accompanying a peer-reviewed article on plant functional traits. • Link: <https://peerj.com/articles/20301/#supp-1>

**Data Structure**

The raw data is an Excel file containing measurements from individual plant samples collected across two major land-use types: • Grouping Variable: Forest type (Categorical: "Conserved" vs. "Disturbed"). • Taxonomic Variable: Plant species and Genus (used as a blocking factor in some analyses). • Trait Variables: The dataset includes a total of 36 quantitative functional traits. These traits were grouped into two categories for analysis: o Morphological Traits (e.g., Specific Leaf Area - SLA, Dry Matter Content - DMC). Ecological Stoichiometric Traits (e.g., concentrations of Carbon, Nitrogen, and Phosphorus in leaf, twig, bark, and stem tissues, and their corresponding stoichiometric ratios like C:N and N:P). 2.

Data Preparation and Cleaning Step 1: Data Import and Formatting The raw data was imported into R using the read_excel() function from the readxl package. The column containing the grouping variable, Forest type, was converted to a factor for use in statistical modeling.

**3. Statistical Analysis**

All analyses were performed to examine the differences in plant functional trait syndromes based on the single grouping factor: Forest Type (Conserved vs. Disturbed).

**Univariate Analysis (ANOVA)**

A one-way Analysis of Variance (ANOVA) was conducted on each of the 36 log-transformed traits, with Forest type as the independent variable. This step determined which individual traits showed a statistically significant difference (p \< 0.05) between the two forest types, serving as the foundation for the subsequent multivariate analyses. The aov() function was used, and the results were summarized in a table.

**Trait Correlation Analysis.**

The relationships between all 36 functional traits were explored using a Pearson correlation matrix. This provided an initial assessment of how traits co-vary, which is essential for understanding functional trade-offs (e.g., the relationship between high Dry Matter Content and low Specific Leaf Area).

**Principal Component Analysis (PCA)**

PCA was performed on the full set of 36 log-transformed traits. The goal of the PCA was dimensionality reduction—to synthesize the high-dimensional trait data into a smaller number of uncorrelated variables (Principal Components or PCs) that capture the majority of the total trait variance. • A Scree Plot was generated to determine the optimal number of PCs to retain based on the "elbow" of the curve. • A Biplot was generated to visualize the trait correlations and sample distribution across the first two most important PCs, revealing the underlying functional axes (e.g., "Structural Investment" vs. "Resource Acquisition").

**Linear Discriminant Analysis (LDA)**

To formally quantify the separation between the two forest types, an LDA was performed. Unlike PCA, which is unsupervised, LDA is a supervised method that explicitly seeks to maximize the separation of the pre-defined groups (Conserved and Disturbed). • The LDA generated a single Discriminant Axis (LD1), which represents the linear combination of traits that best discriminates between the two forest types. • The coefficients of the LD1 equation identified the specific traits (e.g., P concentration, C:P ratio) that were most crucial in differentiating the Conserved and Disturbed samples. • A Density Plot was generated to visualize the distribution of samples along the LD1 axis, allowing for a statistical interpretation of group separation and overlap.

```{r - preparation for data analysis, echo = FALSE}

#Load the data
load("Data/foret.Rdata")
```

## ANOVA  

```{r - ANOVA Species and Genus}

#For loop for ANOVA

responses <- c("leaf.thick", "leaf.area", "spec.leaf.area", "leaf.dry.mat.cont", "leaf.tissue.dens", "twig.dry.mat.cont", "twig.tis.dens", "bark.thick", "bark.dry.mat.cont", "bark.tis.dens", "stem.dry.mat.cont", "stem.tis.dens", "leaf.C.cont", "leaf.N.cont", "leaf.P.cont", "leaf.C.N.ratio", "leaf.C.P.ratio", "leaf.N.P.ratio", "twig.C.cont", "twig.N.cont", "twig.P.cont", "twig.C.N.ratio", "twig.C.P.ratio", "twig.N.P.ratio", "bark.C.cont", "bark.N.cont", "bark.P.cont", "bark.C.N.ratio", "bark.C.P.ratio", "bark.N.P.ratio", "stem.C.content","stem.N.cont", "stem.P.cont", "stem.C.N.ratio", "stem.C.P.ratio", "stem.N.P.ratio")

for (var in responses) {

    # Check first if variable exists in the dataset
    if (!(var %in% names(foret))) { # Using foret_clean is better here
        cat("\nSkipping", var, "-- variable not found in dataset.\n")
        next
    }

    # Model 1: Species + Forest Type
    formula1 <- as.formula(paste(var, "~ Species + For.type"))
  
    model1 <- try(aov(formula1, data = foret), silent = TRUE) 

    # Model 2: Genus + Forest Type
    formula2 <- as.formula(paste(var, "~ genus + For.type"))
    
    model2 <- try(aov(formula2, data = foret), silent = TRUE) 
}

```

```{r - Table for ANOVA}

# List all trait columns (numeric variables)
trait_cols <- names(foret)[sapply(foret, is.numeric)]

# Create empty dataframe
p_table <- data.frame(
  Trait = trait_cols,
  P_value_species = NA,
  P_value_genus = NA
)

# Loop through traits
for (i in seq_along(trait_cols[-1]) +1) {
  trait <- trait_cols[i]
  
# ANOVA for species
  model_species <- aov(foret[[trait]] ~ Species, data = foret)
  p_species <- summary(model_species)[[1]][["Pr(>F)"]][1]
  
# ANOVA for genus
  model_genus <- aov(foret[[trait]] ~ genus, data = foret)
  p_genus <- summary(model_genus)[[1]][["Pr(>F)"]][1]
  
# Save into table
  p_table$P_value_species[i] <- p_species
  p_table$P_value_genus[i] <- p_genus
}
#omit first row of table
p_table <- p_table[-1, ]
row.names(p_table) <- NULL
```

```{r tbl-anovacaption}
#| tbl-cap: "P value ANOVA forest type on Plant morphological traits and ecological stoichiometry, controlling for taxonomic level : species and genus"

library(knitr)
p_table$P_value_species <- as.character(p_table$P_value_species)
p_table$P_value_genus <- as.character(p_table$P_value_genus)

kable(p_table)

```

## ANOVA result interpretation

All variables have significant differences between forest types (see @tbl-anova)

## Correlation

```{r}
#Correlation between the traits

library(GGally)
ggpairs(foret, columns = 4:8,
        aes(color = For.type),
        upper = list(
          continuous = wrap("cor", size = 4)),
        lower = list(
          continuous = wrap("smooth", 
                            alpha = 0.3, size = 0.5))) +
  theme_minimal()

library(GGally)
ggpairs(foret, columns = 9:13,
        aes(color = For.type),
        upper = list(
          continuous = wrap("cor", size = 4)),
        lower = list(
          continuous = wrap("smooth", 
                            alpha = 0.3, size = 0.5))) +
  theme_minimal()

library(GGally)
ggpairs(foret, columns = 14:18,
        aes(color = For.type),
        upper = list(
          continuous = wrap("cor", size = 4)),
        lower = list(
          continuous = wrap("smooth", 
                            alpha = 0.3, size = 0.5))) +
  theme_minimal()

library(GGally)
ggpairs(foret, columns = 19:23,
        aes(color = For.type),
        upper = list(
          continuous = wrap("cor", size = 4)),
        lower = list(
          continuous = wrap("smooth", 
                            alpha = 0.3, size = 0.5))) +
  theme_minimal()

library(GGally)
ggpairs(foret, columns = 24:28,
        aes(color = For.type),
        upper = list(
          continuous = wrap("cor", size = 4)),
        lower = list(
          continuous = wrap("smooth", 
                            alpha = 0.3, size = 0.5))) +
  theme_minimal()

library(GGally)
ggpairs(foret, columns = 29:33,
        aes(color = For.type),
        upper = list(
          continuous = wrap("cor", size = 4)),
        lower = list(
          continuous = wrap("smooth", 
                            alpha = 0.3, size = 0.5))) +
  theme_minimal()

library(GGally)
ggpairs(foret, columns = 34:39,
        aes(color = For.type),
        upper = list(
          continuous = wrap("cor", size = 4)),
        lower = list(
          continuous = wrap("smooth", 
                            alpha = 0.3, size = 0.5))) +
  theme_minimal()

```

## Correlation between traits interpretation

Stem & Leaf nutrient concentrations and tissue density

These figures are pairwise trait‐correlation matrices, split by plant organ (leaf, twig, bark, stem) and forest type (colors: typically conserved vs disturbed forests). Each panel shows:

-   Diagonal: trait distributions (density plots) for the two forest types
-   Lower triangle: scatterplots + fitted trend (separate by forest type)
-   Upper triangle: Pearson correlation (r) with significance (*, **,*** )
-   Colored text = correlations within each forest type

Main patterns

1.  Stem dry matter content ↔ stem tissue density
    -   Strong positive correlation (r ≈ 0.81)
    -   Interpretation: denser stems have more dry matter → structural investment strategy
2.  Leaf N ↔ Leaf P content
    -   Very strong positive correlation (r ≈ 0.82)
    -   Classic leaf economics spectrum: nutrient-rich leaves are coordinated
3.  Leaf C vs N/P
    -   Weak to moderately negative correlations
    -   Carbon-rich leaves tend to be nutrient poorer, reflecting trade-off between structure and metabolism
4.  Forest-type effect
    -   Disturbed forests often show slightly higher nutrient concentrations and wider distributions
    -   Suggests faster-growth strategies under disturbance

Ecological takeaway

Stem traits reflect mechanical strategy, leaf traits reflect metabolic strategy, and these are only weakly coupled.

------------------------------------------------------------------------

Twig & Bark physical traits

Main patterns

1.  Twig dry matter ↔ twig tissue density
    -   Strong positive (r ≈ 0.65)
    -   Denser twigs = more conservative growth
2.  Bark thickness ↔ bark density
    -   Moderate negative correlation
    -   Thick bark tends to be less dense, typical of fire or disturbance protection strategies
3.  Twig traits ↔ bark traits
    -   Generally positive correlations
    -   Indicates whole-plant structural coordination
4.  Forest differences
    -   Disturbed forests show higher variance and steeper slopes
    -   Indicates trait plasticity under stress/disturbance

Ecological takeaway

Twig and bark traits scale together as part of a defensive/structural syndrome, especially important under disturbance.

------------------------------------------------------------------------

Twig nutrient content & stoichiometry

Main patterns

1.  Twig P content ↔ C:P ratio
    -   Strong negative correlation (r ≈ −0.76)
    -   Mathematical + biological: more P → lower C:P → higher nutrient availability
2.  C:N ↔ C:P ratios
    -   Positive correlation
    -   Species with conservative nutrient use are conservative across elements
3.  N:P ratio ↔ C:N / C:P
    -   Negative correlations
    -   Reflects nutrient limitation gradients (N vs P limitation)
4.  Bark C content
    -   Weak correlations → bark is more structural & inert

Ecological takeaway

Twig chemistry strongly reflects nutrient limitation strategies, while bark chemistry is relatively decoupled.

------------------------------------------------------------------------

Leaf stoichiometry & twig nutrients

Main patterns

1.  Leaf C:N ↔ leaf C:P
    -   Very strong positive (r ≈ 0.81)
    -   Strongly coordinated nutrient economics
2.  Leaf N:P ↔ leaf C:P
    -   Positive → confirms P limitation signal
3.  Twig N ↔ leaf N:P
    -   Negative correlation
    -   Higher twig N corresponds to less P limitation in leaves
4.  Forest differences
    -   Disturbed forests cluster toward lower C:N and C:P
    -   Indicates acquisitive nutrient strategy

Ecological takeaway

Leaves are the best indicators of nutrient limitation and productivity strategy.

------------------------------------------------------------------------

Figure 5 — Leaf morphology & economics

Main patterns

1.  Specific leaf area (SLA) ↔ leaf dry matter content
    -   Strong negative correlation (r ≈ −0.67)
    -   Core Leaf Economics Spectrum
2.  Leaf area ↔ SLA
    -   Positive correlation
    -   Large leaves tend to be thinner and resource-acquisitive
3.  Leaf thickness ↔ tissue density
    -   Positive but weaker
    -   Two independent ways to increase structural investment
4.  Forest differences
    -   Disturbed forests show higher SLA and lower tissue density
    -   Indicates fast-growth, short-lived leaves

Ecological takeaway

This figure perfectly captures the fast–slow leaf economics axis.

------------------------------------------------------------------------

Figure 6 — Bark nutrient stoichiometry

Main patterns

1.  Bark N ↔ bark P
    -   Moderate positive correlation
    -   Coordinated nutrient storage
2.  C:N & C:P ratios ↔ N/P contents
    -   Strong negative correlations
    -   Stoichiometry dominated by nutrient dilution
3.  N:P ratio ↔ C:P ratio
    -   Strong positive (r ≈ 0.74)
    -   Indicates P limitation signal even in bark
4.  Forest-type effect
    -   Disturbed forests tend to lower C:N and C:P
    -   More nutrient-rich bark

Ecological takeaway

Bark, though structural, still reflects long-term nutrient strategy.

------------------------------------------------------------------------

Figure 7 — Stem nutrient content & stoichiometry

Main patterns

1.  Stem N ↔ stem P
    -   Strong positive (r ≈ 0.44)
    -   Coordinated nutrient allocation
2.  Stem C:N ↔ stem N content
    -   Strong negative (r ≈ −0.90)
    -   Carbon dilution effect
3.  C:P ↔ N:P ratios
    -   Strong positive
    -   Reflects integrated nutrient limitation
4.  Forest differences
    -   Disturbed forests show higher nutrient contents and lower stoichiometric ratios

Ecological takeaway

Stem chemistry integrates long-term nutrient availability and growth strategy.

------------------------------------------------------------------------

Traits cluster into three coordinated axes:

1\. Structural investment (density, dry matter)\
2. Nutrient economics (N, P, stoichiometry)\
3. Growth strategy (SLA, leaf area)

Disturbed forests consistently show higher nutrient contents, lower C:N and C:P ratios, and higher SLA → faster, more acquisitive strategies.

## Principal Component Analysis

```{r load data}
library(stringr)
#Load the data
load("Data/foret.Rdata")
foret$genus <- word(foret$Species, 1)

```

```{r Setup and Data Preparation.}
ls()
#foret_active <- foret[, c(2:4, 6:10)]
foret_active <- foret[, sapply(foret, is.numeric)]
nrow(foret)
unique(foret$For.type)
unique(foret$Species)


df <- data.frame(
  other_var = c("a", "b", "c"),
  bark.thick = as.character(c(1.5, 2.1, 1.8)),
  bark.dry.mat.cont = as.character(c(0.4, 0.5, 0.3)),
  stem.tis.dens = as.character(c(0.6, 0.7, 0.5)),
  stringsAsFactors = FALSE
)

# A vector containing the names of the columns you want to convert
my_variables <- c("bark.thick", "bark.dry.mat.cont", "bark.tis.dens", "stem.dry.mat.cont", "stem.tis.dens", "bark.C.cont", "bark.N.cont", "bark.P.cont","bark.C.N.ratio", "bark.C.P.ratio", "bark.N.P.ratio", "stem.C.content","stem.N.cont", "stem.P.cont", "stem.C.P.ratio", "stem.N.P.ratio", "stem.C.N.ratio")

for(i in my_variables){
  foret[[i ]] <- as.numeric(foret[[i]])
}

numerical_vars <- names(foret)[sapply(foret, is.numeric)]
foret$bark.thick <- as.numeric(foret$bark.thick)

```

```{r PCA  1 packages for anlysis }

#install.packages("FactoMineR")
#install.packages("factoextra")

library(FactoMineR)
library(factoextra)

res <- PCA(foret_active, scale.unit = TRUE, graph = FALSE)

```

```{r PCA Scree plot and Variance }

graph = FALSE 
res.pca <- PCA(foret_active, scale.unit = TRUE, graph = FALSE)
fviz_eig(res.pca, addlabels = TRUE)
fviz_pca_var(res.pca)
fviz_pca_ind(res.pca,
             geom.ind = "point", # show points only (not text labels)
             col.ind = "cos2",   # Color by quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)
```

The Scree Plot shows the Percentage of Explained Variance for the first 10 dimensions (Principal Components). This plot helps us determine how many components are meaningful to retain for further analysis.

Key Observations:PC1 Dominance (Dimension 1): The first component (Dim 1) is the most important, explaining 24.3% of the total variance in the dataset. PC2 Contribution (Dimension 2): The second component (Dim 2) is also significant, explaining 14.2% of the variance.Cumulative Variance: Together, PC1 and PC2 explain 24.3 + 14.2 = 38.5% of the total variance. The "Elbow": The curve drops steeply from Dim 1 to Dim 2. There is a noticeable "bend" or "elbow" after Dimension 2 or possibly Dimension 3.The drop from Dim 2 (14.2%) to Dim 3 (7.8%) is large. The curve begins to flatten significantly after Dim 3.

The variable plot shows the arrows on the two main components, Dim 1 and Dim 2. Length: Longer arrows indicate variables that are better represented by the two dimensions (i.e., the variables are strongly correlated with PC1 or PC2). Direction (Grouping): Variables pointing in same direction are positively correlated with each other. Variables pointing in opposite directions are negatively correlated.Projection: The closer an arrow is to an axis, the more that variable contributes to that specific component.


```{r  Comparison: Forest Type}
fig.cap="Figure 1:Biplot Grouped by Forest Type"

res.pca <- prcomp(foret[, 3:10],  scale = TRUE)


group_variable <- foret$For.type
fviz_pca_biplot(res.pca, 
                # Color individuals by the 'Species' variable
                habillage = group_variable,
                addEllipses = TRUE, # Optional: Add ellipses around groups
                ellipse.level = 0.95, # Optional: Set ellipse confidence level
                label = "var", # Only label variables (arrows/text)
                repel = TRUE, # Avoid text overlapping
                ggtheme = theme_minimal() # Use a clean ggplot theme
                )
```

The Biplot Grouped by Forest Type Observation Overlap: There is a significant amount of overlap between the two ellipses and the two groups of points. This tells that "Conserved forests" and "Disturbed forests" are not perfectly distinct based.

Separation Along Dim 1 (Horizontal Axis):

The center of the Conserved forests (red ellipse) is shifted slightly to the Left (Negative) side of the Dim 1 axis. The center of the Disturbed forests (blue/teal ellipse) is shifted slightly to the Right (Positive) side of the Dim 1 axis.

Separation Along Dim 2 (Vertical Axis): There is very little separation between the centers of the two groups along the Dim 2 axis. The two groups share a similar range of scores on this second component.

Conserved Forests (Negative Dim 1): These forests tend to be associated with the variables on the left side of the plot, which you identified as having higher Nitrogen (N) Content or Concentration (twig.N.cont, leaf.N.cont, stem.N.cont, etc.). This suggests the trees in conserved forests might have higher overall nutrient concentrations.

Disturbed Forests (Positive Dim 1): These forests tend to be associated with the variables on the right side of the plot, which you identified as representing Structural Carbon-to-Nutrient Ratios and Density/Thickness (bark.dry.mat.cont, leaf.C.P.ratio, etc.). This suggests the trees in disturbed forests might prioritize structural traits or have higher C:P ratios, possibly due to nutrient limitation or different growth strategies in a more stressful/disturbed environment.

```{r  Biplot Grouped by Genus}
fig.cap="Figure 2a: Biplot Grouped by Genus"

res.pca <- prcomp(foret[, 3:10],  scale = TRUE)
genus_counts <- table(foret$genus)
valid_genera <- names(genus_counts[genus_counts >= 3])
foret_filtered <- foret[foret$genus %in% valid_genera, ]
foretnan_filtered <- foret_filtered[, numerical_vars]
res.pca_filtered <- prcomp(foretnan_filtered, scale = TRUE) 
group_variable_filtered <- foret_filtered$genus
 
fviz_pca_biplot(res.pca_filtered, 
                # Color individuals by the filtered genus variable
                habillage = group_variable_filtered,
                addEllipses = TRUE, 
                ellipse.level = 0.95, 
                label = "var", 
                repel = TRUE, 
                # FIX: FORCE a standard, implemented point shape for ALL points
                pointshape = 19, # Use filled circle (19) or open circle (1)
                ggtheme = theme_minimal()
)

res.pca_filtered <- prcomp(foretnan_filtered, scale = TRUE) 
group_variable_filtered <- foret_filtered$genus

fig.cap="Figure 2b: Biplot Grouped by Genus)"
fviz_pca_biplot(res.pca_filtered,                 
                habillage = group_variable_filtered,
                addEllipses = TRUE,                 
                ellipse.level = 0.95,                 
                label = "var",                 
                repel = TRUE,                 
                # FIX: Forces a standard, implemented point shape to fix PCH warnings
                pointshape = 19, 
                ggtheme = theme_minimal()) +   
  ggplot2::theme(legend.position = "none") # Remove the legend
  

```

Interpretation of the Biplot Grouped by Genus The most striking feature is the massive overlap among the ellipses for most of the genera . This indicates that, for the vast majority of genera, their mean structural and nutrient traits fall within a highly similar range in this two-dimensional space. Few Separated Genera (Outliers): Despite the high overlap, a few genera (ellipses) appear to be pushed toward the extremes, suggesting they have unique trait combinations:

Genera on the Far Left (Negative Dim 1): The cluster of ellipses on the negative side of Dim 1 is associated with variables like twig.N.cont, leaf.N.cont, and stem.P.cont. These genera tend to be characterized by high nutrient concentrations (Nitrogen and Phosphorus).

Genera on the Far Right (Positive Dim 1): The cluster of ellipses on the positive side of Dim 1 is associated with variables like twig.C.P.ratio, leaf.C.P.ratio, and bark.dry.mat.cont. These genera tend to be characterized by high structural ratios (high C:P) and high dry matter content/density.

Genera on the Far Bottom (Negative Dim 2): Some groups are pulled strongly toward the bottom of Dim 2, associated with variables like stem.N.ratio and twig.C.N.ratio. These genera exhibit unique characteristics regarding their Nitrogen/Carbon ratios compared to the rest.

## Discriminant Analysis
###  General Discriminant Analysis by Forest

```{r library load}
# Load the MASS package for LDA
library(MASS)
library(ggplot2)
library(dplyr)

```

```{r LDA Analysis}
# Load the necessary library for LDA
library(MASS) 

# --- Data Preparation for LDA ---

# 1. Combine the numerical predictors (foret_active) and the grouping factor (For.type)
lda_data <- foret_active 
lda_data$Group <- foret$For.type

# 2. Filter out rows with NA values (missing data)
# This creates the clean dataset required by the lda() function
lda_data_complete <- na.omit(lda_data)

# 3. Separate the predictors and the factor variable
# All columns except the last (Group) are the predictors
active_vars_complete <- lda_data_complete[, -ncol(lda_data_complete)] 
# The last column is the grouping factor
group_var_complete <- lda_data_complete$Group

# --- Performing the LDA ---

# 4. Perform the LDA
# 'group_var_complete ~ .' predicts the group using all variables in active_vars_complete
lda_result <- lda(group_var_complete ~ ., data = active_vars_complete)

# 5. View the results
print(lda_result)
```

The Coefficients of Linear Discriminants (LD1)

The magnitude and sign of the coefficients tells that:

·Magnitude (Absolute Value): How important the variable is in separating the groups. Larger absolute values mean the variable is a stronger discriminator. The analysis generated one discriminant function, confirming a strong ecological division between the two forest types, primarily driven by a trade-off between Nutrient Acquisition and Structural Investment.

Conserved Forests: These samples are positioned on the negative end of LD1. They exhibit a Nutrient-Acquisition Strategy, characterized by higher by higher absolute N:P concentrations and lower C:P

Disturbed Forests: These samples are positioned on the positive end of LD1. They exhibit a Structural-Investment Strategy, characterized by higher Dry Matter Content and significantly higher C:P

Top Discriminatory Traits (LD1 Coefficients)The coefficients of LD1 indicate the relative contribution and direction of each trait to the separation. The ecological differences are overwhelmingly driven by traits related to Phosphorus cycling and Dry Matter Content (DMC) in structural tissues.The five traits with the largest absolute coefficients, and thus the most influential in the discrimination, are: stem.P.cont, bark.dry.mat.cont,twig.dry.mat.cont, stem.dry.mat.cont, bark.P.cont.

```{r data preparation for visualization}
# 1. Generate the LDA scores (predictions)
# This converts the multidimensional data into scores along the single LD1 axis.
# This assumes 'lda_result' and 'active_vars_complete' are available from the previous step.
lda_scores <- predict(lda_result, active_vars_complete)$x

# 2. Create a data frame for plotting
# We need the LD1 score and the Group variable
lda_plot_data <- data.frame(
  LD1 = lda_scores[, 1],
  Group = group_var_complete
)

# 3. Plot the density curves
fig.cap="Figure 3: Linear Discriminant 1 (LD1) Scores Grouped by Forest Type"


ggplot(lda_plot_data, aes(x = LD1, fill = Group)) +
  # Use geom_density() to show the smooth distribution of scores
  geom_density(alpha = 0.6) +
  # Add a vertical line at 0 for reference (the theoretical decision boundary if priors were equal)
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  labs(
    title = "Separation of Forest Types along Linear Discriminant 1",
    x = "Linear Discriminant 1 (LD1)",
    y = "Density"
  ) +
  # Manually set colors for clarity in the report
  scale_fill_manual(values = c("Conserved forests" = "#FC4E07", "Disturbed forests" = "#00AFBB")) +
  theme_minimal()

```

## Principal Component and Discriminant Analysis Analysis by Genus
```{r loadlibrary}
library(stringr)
library(MASS)
library(ggplot2)
library(dplyr)

```


```{r load data.}
#Load the data
load("Data/foret.Rdata")
foret$genus <- word(foret$Species, 1)


```

```{r Setup and Data Preparation}
# ls()
#foret_active <- foret[, c(2:4, 6:10)]
foret_active <- foret[, sapply(foret, is.numeric)]
# nrow(foret)
# unique(foret$For.type)
# unique(foret$Species)


# df <- data.frame(
#   other_var = c("a", "b", "c"),
#   bark.thick = as.character(c(1.5, 2.1, 1.8)),
#   bark.dry.mat.cont = as.character(c(0.4, 0.5, 0.3)),
#   stem.tis.dens = as.character(c(0.6, 0.7, 0.5)),
#   stringsAsFactors = FALSE
# )

# A vector containing the names of the columns you want to convert
my_variables <- c("bark.thick", "bark.dry.mat.cont", "bark.tis.dens", "stem.dry.mat.cont", "stem.tis.dens", "bark.C.cont", "bark.N.cont", "bark.P.cont","bark.C.N.ratio", "bark.C.P.ratio", "bark.N.P.ratio", "stem.C.content","stem.N.cont", "stem.P.cont", "stem.C.P.ratio", "stem.N.P.ratio", "stem.C.N.ratio")

for(i in my_variables){
  foret[[i ]] <- as.numeric(foret[[i]])
}

numerical_vars <- names(foret)[sapply(foret, is.numeric)]
foret$bark.thick <- as.numeric(foret$bark.thick)
```

```{r PCA 2packages for anlysis }

#install.packages("FactoMineR")
#install.packages("factoextra")

library(FactoMineR)
library(factoextra)

# res <- PCA(foret_active, scale.unit = TRUE, graph = FALSE)

```



```{r Schima  }
genus <- "Schima"

```

```{r multiv}
#| child: _multiv.qmd
```

```{r Acer }
genus <- "Acer"

```

```{r multiv}
#| child: _multiv.qmd
```


```{r Rhododendron}
genus <- "Rhododendron"

```

```{r multiv}
#| child: _multiv.qmd
```


```{r Itea }
genus <- "Itea"

```

```{r multiv}
#| child: _multiv.qmd
```

```{r Phoebe }
genus <- "Phoebe"

```

```{r multiv}
#| child: _multiv.qmd
```

```{r Castanopsis}
genus <- "Castanopsis"
```

```{r multiv}
#| child: _multiv.qmd
```

```{r Cunninghamia}
genus <- "Cunninghamia"
```
```{r multiv}
#| child: _multiv.qmd
```


```{r Choerospondias}
genus <- "Choerospondias"
```
```{r multiv}
#| child: _multiv.qmd
```

```{r Cyclobalanopsis}
genus <- "Cyclobalanopsis"
```
```{r multiv}
#| child: _multiv.qmd
```

```{r Dendrobenthamia}
genus <- "Dendrobenthamia"
```
```{r multiv}
#| child: _multiv.qmd

```

```{r Eurya}
genus <- "Eurya"
```
```{r multiv}
#| child: _multiv.qmd

```

```{r Lindera}
genus <- "Lindera"
```
```{r multiv}
#| child: _multiv.qmd

```

```{r Lithocarpus}
genus <- "Lithocarpus"
```
```{r multiv}
#| child: _multiv.qmd

```

The density plot above projects the scores of all individual forest samples onto the single optimal discriminant axis (LD1). The two distributions clearly illustrate the effectiveness of the measured traits in separating the two forest types. Excellent Separation: The density curves for "Conserved forests" (orange) and "Disturbed forests" (teal) show a high degree of separation with minimal overlap. This confirms that the combined suite of structural and nutritional traits successfully captures the underlying ecological differences between the two groups. Conserved Forests (Orange): The distribution is centered on the negative side of LD1 Mean score approx -1.5. As identified by the negative coefficients in the LDA, this side of the axis is associated with traits like high stem and bark Phosphorus content

Disturbed Forests (Teal): The distribution is centered on the positive side of LD1 (Mean score approx +1.5. This side is associated with positive coefficients for traits like high twig dry matter content

**LDA Result Interpretation**

-   Interpretation of the LDA Density Plot (LD1)

Key Observations: Group Means (Vertical Dashed Lines):

The Conserved forests (red) mean is positioned to the left of zero (in the negative score range).

The Disturbed forests (blue) mean is positioned to the right of zero (in the positive score range).

This confirms the LDA successfully created an axis (LD1) that places the mean of the Conserved group at one end and the mean of the Disturbed group at the other.

-   Overlap and Separation:

High Overlap: There is a substantial region of overlap where the two density curves intersect (roughly between scores -1.0 and 1.5). This indicates that many individual samples, whether Conserved or Disturbed, have very similar trait combinations and would be difficult to classify correctly based on LD1 alone.

Good Separation at Extremes: The LDA does a very good job of identifying the extreme samples. The tails of the distributions (scores below -2.0 and above 2.5) are distinct.

Samples with a very low LD1 score are almost certainly Conserved forests.

Samples with a very high LD1 score are almost certainly Disturbed forests.

Discriminative Traits: The separation is overwhelmingly driven by a trade-off between bark phosphorus content (favoring Conserved forests) and bark tissue density (favoring Disturbed forests)

The `echo: false` option disables the printing of code (only output is displayed).

## Discussion

Separation of Forest Types and the Trait Trade-Off The preliminary analyses consistently demonstrated a significant ecological distinction between the conserved and disturbed forest sites, confirming the Alternate Hypothesis that Plant morphological traits and ecological stoichiometry differ significantly between disturbed and conserved forests

The Analysis of Variance (ANOVA) established that all 36 measured traits, spanning leaf, twig, bark, and stem components, showed significant differences between the two forest types when controlling for taxonomic level (species or genus)

This global difference was then optimally captured by the Linear Discriminant Analysis (LDA), which generated a single axis (LD1) that effectively separated the two groups with minimal overlap at the distribution extremes

The LDA results revealed that the separation along LD1 represents a fundamental ecological trade-off between Nutrient Acquisition and Structural Investment

Conserved Forests (Negative LD1): These samples were overwhelmingly associated with a Nutrient-Acquisition Strategy7. This is characterized by higher absolute concentrations of limiting nutrients, specifically high stem and bark Phosphorus content (stem.P.cont, bark.P.cont)
High nutrient content suggests an environment where resources are efficiently maintained and stored.Disturbed Forests (Positive LD1): These samples clustered on the opposite end of LD1 and were associated with a Structural-Investment Strategy
This strategy is defined by high Dry Matter Content (bark.dry.mat.cont, twig.dry.mat.cont, stem.dry.mat.cont) and high Carbon-to-Phosphorus ratios (leaf.C.P.ratio, twig.C.P.ratio)

The Principal Component Analysis (PCA) and pairwise trait-correlation matrices supported the second Alternate Hypothesis by showing that the traits clustered into three coordinated axes: Structural investment, Nutrient economics, and Growth strategy

In summary, the statistical analyses strongly indicate that the disturbance event has functioned as a powerful environmental filter, successfully selecting for plant species and functional traits that support a fast, resource-acquisitive life history strategy, distinct from the structural and nutrient-conserving strategies prevalent in the conserved forest.

## Conclusions

This preliminary analysis establishes a clear and significant divergence in plant functional traits between the conserved and disturbed forest environments. The discriminant analysis isolated a core ecological trade-off, demonstrating that the disturbed forest community is characterized by a high structural investment strategy (high dry matter content and C:P ratios), while the conserved forest community favors a nutrient acquisition and storage strategy (high absolute N and P concentrations). This functional differentiation confirms the ecological shift imposed by the disturbance, successfully setting the stage for deeper investigation into the underlying drivers of these life history strategies. This aligns with the result from the original paper
